\documentclass[12pt]{lettre}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%~ \usepackage[french]{babel}

\usepackage{graphicx}
\setlength\openingspace{0cm}
\setlength\sigspace{0.2cm}

\setlength{\parskip}{.8em}

\makeatletter
  \def\@texttop{}
\makeatother

\begin{document}

\institut{egce-eng}

\begin{letter}{~}

\nodate
\nolieu

\def\concname{Subject:~}
\conc{Resubmission to Genetics}

\opening{Dear editors, }

We would like to resubmit the attached manuscript, entitled \emph{Gene network simulations provide testable predictions for the
molecular domestication syndrome}, for consideration in Genetics. The previous version of the manuscript (GENETICS-2021-304166) was evaluated positively, but needed to be revised substantially. 

As detailed in the response to the reviewers' comments, we improved the generality of the results by simulating the domestication of four plant species instead of one. We tried to address as thoroughly as possible all remarks raised by the associate editor and by both reviewers. We hope that the revised version has improved substantially the impact of the study and its interest for the readers of Genetics, and that it is now acceptable for publication. 

We followed the Genetics guidelines to format the revised manuscript, including line numbers that were missing in the previous document. We also provide a difference file, highlighting the changes between both versions. 

As indicated in the manuscript, all scripts and programs necessary to reproduce the results and the figures are provided in public (github) directories. We have no conflict of interest to declare. All co-authors have read and validated the submitted version. 

\closing{On behalf of all co-authors, yours sincerely,}

\end{letter}

\end{document}
